
# (PART\*) Introduction {-}

```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# VIDEO Summary of This Course {.unlisted .unnumbered}

# Introduction


## Motivation

How can understanding AI help you be a better leader? 

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/13QC5DTLefV5AIUd_3QPW42D-Zf2Yf6tgrJivIf7jbaQ/edit#slide=id.g1965a5f7f0a_0_44")
```

We think understanding AI is essential for executives. It helps today's leaders make strategic decisions, drive innovation, enhance efficiency, and foster a culture that embraces the transformative power of these technologies. Specifically, AI proficiency can help leaders in the following ways:

1. **Strategic Decision-Making:** Understanding AI and machine learning equips leaders to make informed decisions about integrating these technologies into business strategies, setting their teams up for success when working with AI.

1. **Risk Mitigation:** Familiarity with AI helps leaders assess risks associated with implementing these technologies, ensuring that ethical considerations, data privacy, and potential biases are addressed to mitigate negative consequences. Leaders can also implement more informed policies for their teams.

1. **Efficiency and Experience:** Leaders can explore how AI applications enhance operational efficiency, automate repetitive tasks, and assist employee learning and development, leading to increased productivity and breakthroughs. These improvements can also improve the experience of users or customers your organization serves.

1. **Resource Allocation:** AI resources can be expensive, including in terms of computing resources, subscription services, and/or personnel time. Understanding AI enables leaders to allocate resources effectively, whether in building in-house AI capabilities, partnering with external experts, or investing in AI-driven solutions that align with the organization's mission.

1. **Innovation Leadership:** Leaders can foster a culture of innovation by understanding the transformative potential of AI. Awareness and knowledge can also enable leaders to identify opportunities for innovation, helping their teams match the rapidly evolving technological landscape.

1. **Data-Driven Decision Culture:** Leaders can promote a data-driven decision-making culture within their organizations, using AI insights to inform strategic planning, understand their teams better, and improve other key business functions.

1. **Communication with Tech Teams:** Executives and managers benefit from understanding AI event if they aren't building tech, as it helps them effectively communicate with their technical teams. This can mean more effective collaboration and improved alignment between teams or departments.

<div class = disclaimer>
`r config::get("disclaimer")`
</div>

## Target Audience  

This specialization is intended for executives, decision-makers, and business leaders across industries, including executives in C-suite positions, managers, and directors. Our goal is for these learners to understand the strategic applications of AI and machine learning in driving innovation, improving operations, creating supportive working environments, and gaining an innovative edge. 

We also believe that learning is a life-long process. This specialization is targeted toward those who value continuous learning and want to stay ahead in today's fast-paced technology landscape.

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/13QC5DTLefV5AIUd_3QPW42D-Zf2Yf6tgrJivIf7jbaQ/edit#slide=id.g2a3e68cb5a3_0_16")
```



