
```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# What Is AI

## Machines Can Learn Like Us

Human beings are powerhouses when it comes to pattern recognition and processing [@Mattson2014]. We are constantly observing the world around us, collecting data to learn and make decisions. For example, we might notice a pattern between the amount of traffic on roads in a city, and the time of day. 

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.gcf1264c749_0_140")
```

Much like the human brain, machine learning detects patterns within data. **Machine learning** is at the heart of artificial intelligence, allowing computers to learn and make predictions. In more complex machine learning, computers make millions of calculations, mastering the mapping of inputs (observations) to outputs (predictions). This process mirrors how humans learn through experience. 

<div class = "dictionary"> 
**Machine Learning**: Machine learning is a way for computers to learn from examples and improve their performance over time, resembling how humans learn from experience.
</div>

A machine learning system refines its understanding by continuously updating its parameters based on the feedback received from the provided data. For example, our system might be guessing traffic by time of day, but also judging its accuracy while accounting for other factors, such as whether or not it was a work day, if some workers are on holiday, or how many people live in the city. 

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.g1965a5f7f0a_0_44")
```

The rise of machine learning has been propelled by our ability to collect vast amounts of data and sophisticated types of AI and computing power which we will learn about later.

## The Data Explosion

Let's revisit the traffic example. If you drive once at a specific time of day and observe the traffic around you, you have one data point. You can do this a bunch of times and collect more data. Historically, this is the way data has been collected, and you could manage that data in an Excel Spreadsheet. However, as computer storage and data collection methods have become more sophisticated, our ability to collect data has exploded in scale. It's not hard to imagine that using traffic cameras, dashcams, and car sensors could collect a lot more information than any one person.

Data comes in many shapes and forms. Think about how much text information is freely available on the internet! Using that as input data, artificial intelligence systems can look for patterns of words that typically go together. For example, you're much more likely to see the phrase "cancer is a disease" than "cancer is a computer program". We can see this in action using the AI ChatGPT below.

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.g2a3877ab699_0_79")
```

## AI In Practice

### Amazon Recommendations

Amazon's recommendation engine uses AI algorithms to analyze user behavior and past purchases, providing personalized product recommendations. This enhances the shopping experience, increases customer engagement, and drives sales.

### Another Example

Text here.


## What Is and Is Not AI

It can be tricky to understand what is and isn't AI. Let's look at a few of examples that might seem like AI, but are not.

### Smartphones

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.g2a3877ab699_0_120")
```

The name "smartphone" implies these devices are making decisions and are powered by AI. While there are some features on smartphones that are powered by AI models, like virtual voice assistants and facial recognition, the device as a whole isn't considered AI. 

### Calculators

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.g2a3877ab699_0_279")
```

Many of us use basic calculators, as you might find in Microsoft Excel, every day. AI also makes many calculations. Is it just a scaled=up calculator? No, because AI also makes decisions on new data based on patterns. Traditional calculators can only execute predefined operations.

### Computer Programs

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.g2a3877ab699_0_289")
```

Computers follow set procedures for problem solving and computation. Everyday computers use these proceedures, also called algorithms, to help automate repetitive tasks and save time. However, this isn't generally considered AI. AI systems exhibit the ability to learn, adapt, and handle new inputs for tasks that might be more complicated. 

### Are You Dealing With AI

It can be tricky to determine what kind of tools or systems are AI, especially if you don't come from a computational background. Here are a few questions on which you can reflect to determine if you're dealing with AI or not.

<div class = "reflection"> 
**Does it Learn and Improve Over Time?** : AI systems can learn from feedback and enhance performance without explicit programming.

**Can it Perform Tasks Without Explicit Instructions?** : AI can carry out tasks beyond explicit programming, showcasing adaptability and autonomous decision-making when presented with new data or prompts.

**Does it Exhibit Human-Like Intelligence in its Actions?** : AI mimics aspects of human intelligence, such as problem-solving, pattern recognition, and understanding natural language, even when presented with new data or instructions it hasn't encountered before.
</div>

### DISCUSSION Is It AI

Consider the following examples. Are they examples of AI? Why or why not? Click to expand and see the answer.

<details><summary> <span> **A smartfridge that lets you know when replacement parts are needed**</span></summary><br>

This is not AI. The computer in the fridge is typically programmed to look for specific signs of wear or time passing.

</details>

<details><summary> <span> **Speed cameras on the highway**</span></summary><br>

Speed cameras on highways typically use specialized technology and are not explicitly powered by AI. These cameras are often equipped with radar or lidar sensors for measuring vehicle speed between checkpoints.

While the core functionality of speed cameras relies on sensor technology and predetermined speed thresholds, AI elements may be incorporated in some advanced systems. For example, AI could be used to enhance image recognition accuracy for reading license plates. However, the fundamental operation of speed cameras is rooted in sensor-based speed detection, not AI.
</details>

<details><summary> <span> **Suggested accounts on Instagram**</span></summary><br>

This is considered AI. Social media algorithms, like Instagram's, make recommendations based on user behavior. For example, if you spend a lot of time viewing a page that was recommended, the system interprets that as positive feedback and will make similar recommendations. Typically, these recommendations get better over time as the user generates more user-specific data.

</details>

# VIDEO What Is AI and What It Is Not
