
```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# What Is AI

## Introduction

### Motivation

This course will help you with your understanding of AI, helping you make strategic decision and cultivate a business environment that embraces the benefits of AI, while understanding its limitations and risks.

### Target Audience

This course is targeted toward industry and non-profit leaders and decision makers.

### Curriculum Summary

In this course, we'll learn about what Artificial intelligence is, and what it isn't. We'll also learn the basics of how it works, learn about different types of AI, and set some ground rules for minimizing the harms and maximizing the benefits of AI.

## AI In Practice

### Amazon Recommendations

Amazon's recommendation engine uses AI algorithms to analyze user behavior and past purchases, providing personalized product recommendations. This enhances the shopping experience, increases customer engagement, and drives sales.

### Another Example

Text here.

## Machines Can Learn Like Us

Human beings are powerhouses when it comes to pattern recognition and processing [@Mattson2014]. We are constantly observing the world around us, collecting data to learn and make decisions. For example, we might notice a pattern between the amount of traffic on roads in a city, and the time of day. 

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.gcf1264c749_0_140")
```

Much like the human brain, machine learning detects patterns within data. **Machine learning** is at the heart of artificial intelligence, allowing computers to learn and make predictions. In more complex machine learning, computers make millions of calculations, mastering the mapping of inputs (observations) to outputs (predictions). This process mirrors how humans learn through experience. 

<div class = "dictionary"> 
**Machine Learning**: Machine learning is a way for computers to learn from examples and improve their performance over time, resembling how humans learn from experience.
</div>

A machine learning system refines its understanding by continuously updating its parameters based on the feedback received from the provided data. For example, our system might be guessing traffic by time of day, but also judging its accuracy while accounting for other factors, such as whether or not it was a work day, if some workers are on holiday, or how many people live in the city. 

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.g1965a5f7f0a_0_44")
```

The rise of machine learning has been propelled by our ability to collect vast amounts of data and sophisticated types of AI and computing power which we will learn about later.

## The Data Explosion

Let's revisit the traffic example. If you drive once at a specific time of day and observe the traffic around you, you have one data point. You can do this a bunch of times and collect more data. Historically, this is the way data has been collected, and you could manage that data in an Excel Spreadsheet. However, as computer storage and data collection methods have become more sophisticated, our ability to collect data has exploded in scale. It's not hard to imagine that using traffic cameras, dashcams, and car sensors could collect a lot more information than any one person.

Data comes in many shapes and forms. Think about how much text information is freely available on the internet! Using that as input data, artificial intelligence systems can look for patterns of words that typically go together. For example, you're much more likely to see the phrase "cancer is a disease" than "cancer is a computer program". We can see this in action using the AI ChatGPT below.

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.g2a3877ab699_0_79")
```

## What Is and Is Not AI

It can be tricky to understand what is and isn't AI. Let's look at a few of examples.

### Smartphones

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.g2a3877ab699_0_120")
```

The name "smartphone" implies these devices are making decisions and are powered by AI. While there are some features on smartphones that are powered by AI models, like virtual voice assistants and facial recognition, the device as a whole isn't considered AI. 

### Calculators

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.g2a3877ab699_0_279")
```

Many of us use basic calculators, as you might find in Microsoft Excel, every day. AI also makes many calculations. Is it just a scaled=up calculator? No, because AI also makes decisions on new data based on patterns. Traditional calculators can only execute predefined operations.

### Computer Programs

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.g2a3877ab699_0_289")
```

Computers follow set procedures for problem solving and computation. Everyday computers use these proceedures, also called algorithms, to help automate repetitive tasks and save time. However, this isn't generally considered AI. AI systems exhibit the ability to learn, adapt, and handle new inputs for tasks that might be more complicated. 

### DISCUSSION Is It AI

It can be tricky to determine what kind of tools or systems are AI, especially if you don't come from a computational background. Here are a few questions you can reflect on to determine if 

# VIDEO What Is AI and What It Is Not
