
```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# VIDEO What Is AI

TODO: Slides here: https://docs.google.com/presentation/d/1-Mm-Vym3xdtB8xLRHR24jNCLSbNgFHw6N62iJrYe63c/edit#slide=id.g2a162964683_1_0

# What Is AI

When discussing artificial intelligence, people often envision humanoid robots, prompting concerns about their ability to outsmart us. The notion of robots passing tests that blur the line between human and machine, often depicted in science fiction, adds to these worries, particularly when considering the potential for AI systems to act in self-interest and make decisions independently.

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1-Mm-Vym3xdtB8xLRHR24jNCLSbNgFHw6N62iJrYe63c/edit#slide=id.g2a68fa11a90_0_0")
```

Scientists are still working toward computers with full human problem solving and cognition, or **artificial general intelligence**. We aren't there yet. Currently, artificial intelligence systems are optimized to perform a specific task well, but not for general, multi-purpose tasks. For example, the AI application for recognizing voices can not be directly applied to drive cars, and vice versa. Similarly, a language translation app could not recognize images, and vice versa.


```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1-Mm-Vym3xdtB8xLRHR24jNCLSbNgFHw6N62iJrYe63c/edit#slide=id.gcf1264c749_0_130")
```


## AI In Practice

The following are case studies that can help us conceptualize AI in the real world.

### Amazon Recommendations

Amazon's recommendation engine uses AI algorithms to analyze user behavior and past purchases, providing personalized product recommendations. This enhances the shopping experience, increases customer engagement, and drives sales.

TODO: Text here.

### Financial Forecasting

TODO: Text here.

## What Is and Is Not AI

It can be tricky to understand what is and isn't AI. Let's look at a few of examples that might seem like AI, but are not.

### Smartphones

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.g2a3877ab699_0_120")
```

The name "smartphone" implies these devices are making decisions and are powered by AI. While there are some features on smartphones that are powered by AI models, like virtual voice assistants and facial recognition, the device as a whole isn't considered AI. 

### Calculators

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.g2a3877ab699_0_279")
```

Many of us use basic calculators, as you might find in Microsoft Excel, every day. AI also makes many calculations. Is it just a scaled=up calculator? No, because AI also makes decisions on new data based on patterns. Traditional calculators can only execute predefined operations.

### Computer Programs

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.g2a3877ab699_0_289")
```

Computers follow set procedures for problem solving and computation. Everyday computers use these procedures, also called algorithms, to help automate repetitive tasks and save time. However, this isn't generally considered AI. AI systems exhibit the ability to learn, adapt, and handle new inputs for tasks that might be more complicated. 

### Are You Dealing With AI

It can be tricky to determine what kind of tools or systems are AI, especially if you don't come from a computational background. Here are a few questions on which you can reflect to determine if you're dealing with AI or not.

<div class = "reflection"> 
**Does it Learn and Improve Over Time?** : AI systems can learn from feedback and enhance performance without explicit programming.

**Can it Perform Tasks Without Explicit Instructions?** : AI can carry out tasks beyond explicit programming, showcasing adaptability and autonomous decision-making when presented with new data or prompts.

**Does it Exhibit Human-Like Intelligence in its Actions?** : AI mimics aspects of human intelligence, such as problem-solving, pattern recognition, and understanding natural language, even when presented with new data or instructions it hasn't encountered before.
</div>

### DISCUSSION Is It AI

Consider the following examples. Are they examples of AI? Why or why not? Click to expand and see the answer.

<details><summary> <span> **A smartfridge that lets you know when replacement parts are needed**</span></summary><br>

This is not AI. The computer in the fridge is typically programmed to look for specific signs of wear or time passing.

</details>

<details><summary> <span> **Speed cameras on the highway**</span></summary><br>

Speed cameras on highways typically use specialized technology and are not explicitly powered by AI. These cameras are often equipped with radar sensors for measuring vehicle speed between checkpoints.

While the core functionality of speed cameras relies on sensor technology and predetermined speed thresholds, AI elements may be incorporated in some advanced systems. For example, AI could be used to enhance image recognition accuracy for reading license plates. However, the fundamental operation of speed cameras is rooted in sensor-based speed detection, not AI.
</details>

<details><summary> <span> **Suggested accounts on Instagram**</span></summary><br>

This is considered AI. Social media algorithms, like Instagram's, make recommendations based on user behavior. For example, if you spend a lot of time viewing a page that was recommended, the system interprets that as positive feedback and will make similar recommendations. Typically, these recommendations get better over time as the user generates more user-specific data.

</details>
