
```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# VIDEO How AI Works

# How AI Works

Let's briefly revisit our definition of AI: it must have data, training via an algorithm, and an interface. How do each of these work? We'll explore below.

## The Data Explosion

Let's say we're driving a car or taking public transportation in a city. We might notice a pattern between the amount of traffic on roads, and the time of day. If you commute once at a specific time of day and observe the traffic around you, you have one data point. You can do this a bunch of times and collect more data. 

Historically, this is the way data has been collected, and you could manage that data in an Excel Spreadsheet. However, as computer storage has become cheaper and data collection methods have become more sophisticated, our ability to access data has exploded in scale. It's not hard to imagine that using traffic cameras, dashcams, and car sensors could collect a lot more information than any one person.

Think about how much text information is freely available on the internet! Treating that as input data, AI systems can look for patterns of words that typically go together. For example, you're much more likely to see the phrase "cancer is a disease" than "cancer is a computer program".

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.g2a3877ab699_0_79")
```

### What Is Data

Data comes in many shapes and forms. Data can be **structured**, such as a spreadsheet of times and traffic volume or counts of viral particles in different patients. Data can also be **unstructured**, such as might be found in social media text or genome sequence data.

Other kinds of data can be collected and used to train algorithms. These might include survey data collected directly from consumers, medical data collected in a healthcare setting, purchase or transaction tracking, and online tracking of your time on certain web pages [@Cote2022].

<div class = "ethics"> 
It is **essential** that you and your team think critically about data sources. Many companies releasing generative AI systems have come under fire for training these systems on data that doesn't belong to them [@Walsh2023]. Individual people also have a right to data privacy. No personal data should be used without permission, even if that data could be interesting or useful.  
</div>

## Machines Can Learn Like Us

Human beings are powerhouses when it comes to pattern recognition and processing [@Mattson2014]. We are constantly observing the world around us, collecting data to learn and make decisions. For example, we might notice a pattern between the amount of traffic on roads in a city, and the time of day. 

```{r, eval=FALSE, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.gcf1264c749_0_140")
```



Much like the human brain, machine learning detects patterns within data. **Machine learning** is at the heart of artificial intelligence, allowing computers to learn and make predictions. In more complex machine learning, computers make millions of calculations, mastering the mapping of inputs (observations) to outputs (predictions). This process mirrors how humans learn through experience. 

<div class = "dictionary"> 
**Machine Learning**: Machine learning is a way for computers to learn from examples and improve their performance over time, resembling how humans learn from experience.
</div>

A machine learning system refines its understanding by continuously updating its parameters based on the feedback received from the provided data. For example, our system might be guessing traffic by time of day, but also judging its accuracy while accounting for other factors, such as whether or not it was a work day, if some workers are on holiday, or how many people live in the city. 

```{r, echo=FALSE, eval=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.g1965a5f7f0a_0_44")
```

The rise of machine learning has been propelled by our ability to collect vast amounts of data and sophisticated types of AI and computing power.




<!-- ## Motivation -->


<!-- ## Target Audience   -->

<!-- The course is intended for ... -->

<!-- ## Curriculum   -->

<!-- The course covers... -->

