
```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# VIDEO How AI Works

# How AI Works

Let's briefly revisit our definition of AI: it must have data, training via an algorithm, and an interface. Let's dive into each of these in more detail below.

## Early Warning for Skin Cancer

Each year in the United States, 6.1 million adults are treated for skin cancer (basal cell and squamous cell carcinomas), totaling nearly $10 billion in costs [@CDC2023]. It is one of the most common forms of cancer in the United States, and mortality from skin cancer is a real concern. Fortunately, early detection through regular screening can increase survival rates to over 95% [@Melarkode2023]. Cost and accessibility of screening providers, however, means that many people aren't getting the preventative care they need.

Increasingly, AI is being used to flag potential skin cancer. AI focused on skin cancer detection could be used by would-be patients to motivate them to seek a professional opinion, or by clinicians to validate their findings or help with continuous learning.

1. **Data**: Images of skin

1. **Algorithm**: Detection of possible skin cancer

1. **Interface**: Web portal or app where you can submit a new picture

## Collecting Datapoints

Let's say a clinician, *Dr. J*, is learning how to screen for skin cancer. Dr. J diagnoses their first instance of skin cancer, meaning they now have one internal reference point. Dr. J could make future diagnoses based on this one data point, but it might not be very accurate. Over time, as Dr. J does more screenings of skin with and without cancer, they will get a better and better idea of what skin cancer looks like. Many of us refer to our on-the-job learning as "training".

AI works in a similar way. As more data is provided, AI will typically get closer to finding the patterns in which we are interested. In order to train an AI algorithm to detect possible skin cancer, we'll first want to gather as many pictures of normal and cancerous skin as we can. This is our **raw data** [@Leek2017].

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.g263e06ef889_36_153")
```

### What Is Data

In our skin cancer screening example, our data is all of the information stored in an image. However, data comes in many shapes and forms. Data can be **structured**, such as a spreadsheet of the time of day plus traffic volume or counts of viral particles in different patients. Data can also be **unstructured**, such as might be found in social media text or genome sequence data. 

Other kinds of data can be collected and used to train algorithms. These might include survey data collected directly from consumers, medical data collected in a healthcare setting, purchase or transaction tracking, and online tracking of your time on certain web pages [@Cote2022].

Quantity *and* quality of data are very important. More data makes it easier to detect and account for minor differences among observations. However, that shouldn't come at the cost of quality. It is sometimes better to have fewer, high resolution or high quality images in our dataset than many images that are blurry, discolored, or in other ways questionable.

Large Language Models (LLMs), which we will cover later, are great examples of high quantity and quality of data. Think about how much text information is freely available on the internet! Treating that as input data, AI systems can look for patterns of words that typically go together. For example, you're much more likely to see the phrase "cancer is a disease" than "cancer is a computer program". Many LLMs are trained on sources like [Wikipedia](https://www.wikipedia.org/), which are typically grammatically sound and informative, leading to higher quality output.

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.g2a3877ab699_0_79")
```

<div class = "ethics"> 
It is **essential** that you and your team think critically about data sources. Many companies releasing generative AI systems have come under fire for training these systems on data that doesn't belong to them [@Walsh2023]. Individual people also have a right to data privacy. No personal data should be used without permission, even if that data could be interesting or useful.  
</div>

### Preparing the Data

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.g263e06ef889_36_318")
```




## Machines Can Learn Like Us

Human beings are powerhouses when it comes to pattern recognition and processing [@Mattson2014]. We are constantly observing the world around us, collecting data to learn and make decisions. For example, we might notice a pattern between the amount of traffic on roads in a city, and the time of day. 

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.gcf1264c749_0_140")
```

Much like the human brain, machine learning detects patterns within data. **Machine learning** is at the heart of artificial intelligence, allowing computers to learn and make predictions. In more complex machine learning, computers make millions of calculations, mastering the mapping of inputs (observations) to outputs (predictions). This process mirrors how humans learn through experience. 

<div class = "dictionary"> 
**Machine Learning**: Machine learning is a way for computers to learn from examples and improve their performance over time, resembling how humans learn from experience.
</div>

A machine learning system refines its understanding by continuously updating its parameters based on the feedback received from the provided data. For example, our system might be guessing traffic by time of day, but also judging its accuracy while accounting for other factors, such as whether or not it was a work day, if some workers are on holiday, or how many people live in the city. 

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.g1965a5f7f0a_0_44")
```

The rise of machine learning has been propelled by our ability to collect vast amounts of data and sophisticated types of AI and computing power.
