
```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# VIDEO How AI Works

# How AI Works

## Machines Can Learn Like Us

Human beings are powerhouses when it comes to pattern recognition and processing [@Mattson2014]. We are constantly observing the world around us, collecting data to learn and make decisions. For example, we might notice a pattern between the amount of traffic on roads in a city, and the time of day. 

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.gcf1264c749_0_140")
```

Much like the human brain, machine learning detects patterns within data. **Machine learning** is at the heart of artificial intelligence, allowing computers to learn and make predictions. In more complex machine learning, computers make millions of calculations, mastering the mapping of inputs (observations) to outputs (predictions). This process mirrors how humans learn through experience. 

<div class = "dictionary"> 
**Machine Learning**: Machine learning is a way for computers to learn from examples and improve their performance over time, resembling how humans learn from experience.
</div>

A machine learning system refines its understanding by continuously updating its parameters based on the feedback received from the provided data. For example, our system might be guessing traffic by time of day, but also judging its accuracy while accounting for other factors, such as whether or not it was a work day, if some workers are on holiday, or how many people live in the city. 

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.g1965a5f7f0a_0_44")
```

The rise of machine learning has been propelled by our ability to collect vast amounts of data and sophisticated types of AI and computing power which we will learn about later.

## The Data Explosion

Let's revisit the traffic example. If you drive once at a specific time of day and observe the traffic around you, you have one data point. You can do this a bunch of times and collect more data. Historically, this is the way data has been collected, and you could manage that data in an Excel Spreadsheet. However, as computer storage and data collection methods have become more sophisticated, our ability to collect data has exploded in scale. It's not hard to imagine that using traffic cameras, dashcams, and car sensors could collect a lot more information than any one person.

Data comes in many shapes and forms. Think about how much text information is freely available on the internet! Using that as input data, artificial intelligence systems can look for patterns of words that typically go together. For example, you're much more likely to see the phrase "cancer is a disease" than "cancer is a computer program". We can see this in action using the AI ChatGPT below.

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1OydUyEv1uEzn8ONPznxH1mGd4VHC9n88_aUGqkHJX4I/edit#slide=id.g2a3877ab699_0_79")
```


<!-- ## Motivation -->


<!-- ## Target Audience   -->

<!-- The course is intended for ... -->

<!-- ## Curriculum   -->

<!-- The course covers... -->

