
```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# VIDEO Different Types of AI

# Types of AI

We've learned a bit about how AI works. However there are many different types of AI with different combinations of data, algorithms, and interfaces. There are also general terms that are important to know. Let's explore some of these below.

## Machine Learning 

**Machine learning** is broad concept describing how computers to learn from data. It includes traditional methods like decision trees and linear regression, as well as more modern approaches such as deep learning. It involves training models on labeled data to make predictions or uncover patterns or grouping of data. Machine learning is often the "algorithm" part of our data - algorithm - interface framework.

## Neural Networks

**Neural networks** are a specific class of algorithms within the broader field of machine learning. They organize data into layers, including an input layer for data input and an output layer for results, with intermediate layers in between. These layers help neural networks understand hierarchical patterns in data. 

The connections between nodes have weights that the network learns during training. The network can then adjust these weights to minimize errors in predictions. Neural networks often require large amounts of labeled data for training, and their performance may continue to improve with more data.

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1UiYOR_4a68524XsCv-f950n_CfbyNJVez2KdAjq2ltU/edit#slide=id.g2a694e3cce9_0_0")
```
