
```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# VIDEO Different Types of AI

# Demystifying Types of AI

We've learned a bit about how AI works. However there are many different types of AI with different combinations of data, algorithms, and interfaces. There are also general terms that are important to know. Let's explore some of these below.

## Machine Learning 

**Machine learning** is broad concept describing how computers to learn from data. It includes traditional methods like decision trees and linear regression, as well as more modern approaches such as deep learning. It involves training models on labeled data to make predictions or uncover patterns or grouping of data. Machine learning is often the "algorithm" part of our data - algorithm - interface framework.

## Neural Networks

**Neural networks** are a specific class of algorithms within machine learning. Neural networks mimic the way data is transferred between neurons in the brain.

Neural networks organize data into layers, starting with an "input layer" of raw data. Data is then transferred to the next layer, sometimes called a "hidden" layer. The hidden layer combines the raw data in many ways to create levels of abstraction. You can think of an image that is very pixelated becoming more clear. Finally, results are produced in an "output layer".

Neural networks often require large amounts of labeled data for training, and their performance may continue to improve with more data.

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/1UiYOR_4a68524XsCv-f950n_CfbyNJVez2KdAjq2ltU/edit#slide=id.g2a694e3cce9_0_0")
```

## Deep Learning

**Deep learning** refers to neural networks with multiple intermediate "hidden" layers. A neural network with 2+ hidden layers could be considered a deep learning system [@ibm2023]. 