
```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

# VIDEO Real Life Possibilities

You can find the Google Slides for this video [here](https://docs.google.com/presentation/d/16FlAoE9-kTinf4ErYYtDs5Ge2thneUtQUawem_eaHtc/edit?usp=sharing).

# What AI Makes Possible

<!-- What is possible with AI? What’s still fantasy? -->

AI is opening up many possible pathways in many different fields. Let's explore several of the general ways in which AI can be used today.

## Advancements in Text Mining

**Text mining** is the process of extracting meaningful insights, patterns, and knowledge from unstructured textual data. This data could include articles, documents, emails, social media posts, business records, policy records, and more. This data is digested by a computer into a structured format for analysis, allowing for the discovery of hidden patterns, relationships, and / or summary information.

Historically, mining for relevant text had to follow rule-based or statistical methods that required a lot of human oversight. Generative AI has led to many advancements in text mining. Some of these include:

1. **Contextual Understanding**: Generative AI, especially transformer models, has improved contextual understanding in text mining. AI can consider relationships between words in a sentence more effectively using the self-attention mechanisms available as part of transformers. This results in more accurate extraction of context-dependent information.

1. **Text Completion and Generation**: Generative AI allows for the completion of partial or missing text. In text mining, this capability is useful for handling incomplete or noisy data, improving the overall quality of mined information.

1. **Domain-Specific Language Generation**: Generative AI can be fine-tuned for domain-specific language generation. This is particularly beneficial in industries where specialized terminology or jargon is prevalent. By training generative models on domain-specific data, text mining models can better adapt to the nuances of the industry or discipline in general.

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/16FlAoE9-kTinf4ErYYtDs5Ge2thneUtQUawem_eaHtc/edit#slide=id.g2641ae2cc7f_0_111")
```

### Examples

Text mining can be used for:

1. Mining clinical patient notes to identify patients with similar symptoms

1. Mining extensive technical or financial documents to identify relevant sections more easily 

A simplified example of text mining can be seen here with [ChatGPT](https://chat.openai.com/). We will provide some information from the [Wikipedia article for the city of Seattle](https://en.wikipedia.org/wiki/Seattle) and ask the large language model to extract only relevant statistics related to population growth.

> Seattle (/siˈætəl/ ⓘ see-AT-əl) is a seaport city on the West Coast of the United States. It is the seat of King County, Washington. With a 2022 population of 749,256[10] it is the most populous city in both the state of Washington and the Pacific Northwest region of North America, and the 18th most populous city in the United States. The Seattle metropolitan area's population is 4.02 million, making it the 15th-largest in the United States.[11] Its growth rate of 21.1% between 2010 and 2020 made it one of the country's fastest-growing large cities.[12]

> Seattle is situated on an isthmus between Puget Sound (an inlet of the Pacific Ocean) and Lake Washington. It is the northernmost major city in the United States, located about 100 miles (160 km) south of the Canadian border. A major gateway for trade with East Asia, the Port of Seattle is the fourth-largest port in North America in terms of container handling as of 2021.[13]

> The Seattle area has been inhabited by Native Americans (such as the Duwamish, who had at least 17 villages around Elliot Bay) for at least 4,000 years before the first permanent European settlers.[14] Arthur A. Denny and his group of travelers, subsequently known as the Denny Party, arrived from Illinois via Portland, Oregon, on the schooner Exact at Alki Point on November 13, 1851.[15] The settlement was moved to the eastern shore of Elliott Bay in 1852 and named "Seattle" in honor of Chief Seattle, a prominent 19th-century leader of the local Duwamish and Suquamish tribes. Seattle currently has high populations of Native Americans alongside Americans with strong Asian, African, European, and Scandinavian ancestry, and hosts the fifth-largest LGBT community in the U.S.[16]

Here is the response:

> Certainly! Here are the relevant statistics extracted from the provided passage related to population growth:

> 1. Population Growth Rate (2010-2020): The growth rate of Seattle between 2010 and 2020 was 21.1%, making it one of the country's fastest-growing large cities.[12]

> 2. Current Population (2022): The population of Seattle in 2022 is reported to be 749,256.[10]

> 3. Metropolitan Area Population: The population of the Seattle metropolitan area is 4.02 million, making it the 15th-largest in the United States.[11]

> These statistics highlight the significant population growth of Seattle over the specified time period, positioning it as one of the fastest-growing large cities in the United States.

In reality, you might be providing a much larger volume of data or be looking for more nuanced information.

```{r, echo=FALSE, fig.alt='CAPTION HERE', out.width = '100%', fig.align = 'center'}
ottrpal::include_slide("https://docs.google.com/presentation/d/16FlAoE9-kTinf4ErYYtDs5Ge2thneUtQUawem_eaHtc/edit#slide=id.g2641ae2cc7f_0_140")
```

## Modifying and Generating Text

Generative AI in the form of Large language Models like [ChatGPT](https://chat.openai.com/) and [Claude](https://poe.com/Claude-instant) can be especially useful for modifying and generating text. This could mean changing the tone of a piece of writing, creating summaries based on text data input, or making writing more understandable to different audiences.

### Examples

Here are some examples of using AI to modify or generate text:

1. Altering an email to be more empathetic

1. Summarizing multiple files within a project to make a project summary - For example, a README file across many code scripts for software development

1. Summarizing legal documents in a way that is readable to broader audiences with less familiarity with legal jargon

Let's see some of these in practice. Take for example this email to a colleague that is really frustrating you right now.

> 


# VIDEO What Is Possible

# VIDEO What Is NOT Possible

<div class = disclaimer>
`r config::get("disclaimer")`
</div>
