
# (PART\*) Determining AI Needs {-}

```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

<!-- # VIDEO Introduction to Determining AI Needs Video -->

<!-- This video introduces the 'Determining AI Needs' course. -->

<!-- <!-- <iframe width="560" height="315" src="https://www.youtube-nocookie.com/embed/ID" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe> -->  -->

<!-- You can view and download the Google Slides [here](https://docs.google.com/presentation/d/1COHDxEwy9GwXAgUJLBqjDjWm-lqdKy2n3Qds4ivM4UA/edit). -->


# Introduction to Determining AI Needs

<div class = disclaimer>
`r config::get("disclaimer")`
</div>


## Motivation

There are an ever increasing number of options, strategies and solutions for integrating AI solutions into a project. It can feel overwhelming to understand what these options entail let alone understand how to decide what solution best fits a use case.

In this course we aim to give individuals the basic info they need to make basic plans for integrating AI tools into their project.

## Target Audience  

The course is intended for individuals who have an AI related project in mind or think that they might need to incorporate AI into their project. They are likely the leader who is guiding others in an AI related project and not necessarily the person who will write code or carry out the technical aspects of the project.

```{r for_individuals_who, fig.align='center', out.width="100%", echo = FALSE, fig.alt= "This course is designed for leaders, executives, and decision-makers shaping strategy and driving innovation for use or development of AI who need to learn if premade AI solutions will work for you or custom solutions are needed and need to learn now to evaluate AI tools and resources."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1Kg_uBagr7rdUX4okVbaxpV0QMdO4nm8o2-KQY7AECeE/edit?slide=id.g34ec17a9f7c_0_0#slide=id.g34ec17a9f7c_0_0")
```

## Curriculum  

```{r topics_covered, echo=FALSE, fig.alt='Pathway of more specific topics covered in this course: In this course, you’ll learn the components of good AI systems, how to determine if you can use an out of the box existing tool or if you need to make a custom tool from scratch, considerations for customizing the knowledge base, security methods, and interface, and how to evaluate your own custom AI tool.', out.width = '100%', warning=FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1Kg_uBagr7rdUX4okVbaxpV0QMdO4nm8o2-KQY7AECeE/edit?slide=id.g34ec17a9f7c_0_8#slide=id.g34ec17a9f7c_0_8")
```


**What this course covers:**

- What are the practical aspects of AI that need to be understood before endeavoring on an AI project?
- What makes an AI model good?
- How do you determine what kinds of custom AI solutions your project needs if any?
- What aspects of your resources and your project should you consider when evaluating AI strategies?
- What would better suite your needs an "out of the box" AI product or building an AI model solution "from scratch"?
- Examples of currently existing AI solutions that may suit an individual's AI needs.

**What this course does NOT cover:**

- This is NOT a comprehensive survey of the AI tools and products in existence. Even if it was comprehensive at this time, there are new tools and developments constantly arriving. We merely give examples of solutions that show a *possible* AI solution. There may be competitors or similar solutions out there that would even better fit a project's needs.
- This does NOT cover in depth aspects of algorithms, statistics or mathematics behind AI algorithms -- these are numerous and not always necessary to understand in fine detail for making decisions about projects.
- This does NOT cover how to complete or write code for an AI project. This is not a tutorial for building an AI tool. Instead we merely give strategies you could employ but we do not give details on how you might employ them. There are too many ways that AI tools may be built -- this is outside the scope of this course.




## Learning Objectives

```{r learning_objectives, echo=FALSE, fig.alt = 'The overall course learning objectives are: By completing this course, learners will be able to Understand practical aspects of AI components, Identify what makes an AI model good, Determine when custom AI solutions might be needed for a project or when existing options might work well, and Identify examples of currently existing AI solutions', out.width = '100%', warning=FALSE}
ottrpal::include_slide("https://docs.google.com/presentation/d/1Kg_uBagr7rdUX4okVbaxpV0QMdO4nm8o2-KQY7AECeE/edit?slide=id.g34ec17a9f7c_0_37#slide=id.g34ec17a9f7c_0_37")
```

